---
title: "Elite Identification in Latin America"
subtitle: "Networks, Text Mining & NLP"
author: "Matías Deneken"
date: "April 2026"
format:
  revealjs:
    theme: dark
    slide-number: true
    transition: fade
    background-transition: fade
    footer: "8th Annual COMPTEXT Conference 2026"
    css: comptext-theme.css
---

## Motivation

Identifying political and economic elites in Latin America is central to the study of power and inequality.

- **Structured data** (biographies, kinship, offices) is scattered across languages and sources.
- **Text mining** and **NLP** help extract entities, relations, and roles from unstructured text.
- **Network analysis** turns those relations into graphs: families, institutions, cross-country ties.

::: {.fragment}
Combining NLP + networks yields a computational approach to elite identification and comparison across countries.
:::

## Pipeline Overview

:::: {.columns}

::: {.column width="50%"}
**Data & text**

- Wikipedia biographies (multilingual).
- Infoboxes and structured fields.
- Person–person relations (family, office).
:::

::: {.column width="50%"}
**Methods**

- Entity extraction and normalization.
- Relation extraction (NLP / rules).
- Graph construction and centrality.
- Cross-country linkage.
:::

::::

## From Text to Networks

1. **Text** → entities (persons, roles, places) via parsing and (optionally) NER.
2. **Relations** → edges (marriage, parent–child, shared office, diplomatic links).
3. **Graph** → nodes = persons/families; edges = relations; metrics = degree, betweenness, communities.

::: {.fragment}
The same pipeline supports **within-country** elite networks and **cross-country** family/diplomatic ties.
:::

## Example: Within-Country Networks

```{r}
#| echo: false
#| fig-align: center
#| fig-cap: "Red de personas relacionadas por país (ejemplo Chile)."
knitr::include_graphics(
  if (file.exists("../outputs/figures/exploracion_redes_pais/red_chile.png"))
    "../outputs/figures/exploracion_redes_pais/red_chile.png"
  else
    "https://placehold.co/800x500/1a2332/5eb3d6?text=Red+por+país"
)
```

Families and key connectors emerge from degree and betweenness; layout highlights clusters.

## Example: Cross-Country Ties

```{r}
#| echo: false
#| fig-align: center
#| fig-cap: "Uniones familiares y vínculos diplomáticos entre países."
knitr::include_graphics(
  if (file.exists("../outputs/figures/exploracion_union_entre_paises.png"))
    "../outputs/figures/exploracion_union_entre_paises.png"
  else
    "https://placehold.co/800x450/1a2332/8b5cf6?text=Red+entre+países"
)
```

Nodes = countries; edges = family unions or diplomatic/professional links inferred from text and metadata.

## Text Mining in the Workflow

- **Structured fields**: infoboxes, categories, “personas relacionadas” (Wikipedia).
- **Unstructured**: biography text for nationality, residence, offices, death place.
- **Inference**: cross-country links from `nacionalidad`, `lugar_nacimiento`, `cargos_politicos`, `residencia`.
- **Optional**: NER, relation extraction, or LLMs for roles and ties when structure is missing.

::: {.fragment}
Text mining supports both **node attributes** and **edge creation** in the elite network.
:::

## Tools & Stack

| Layer        | Role                          |
|-------------|--------------------------------|
| **Data**    | Wikipedia API, CSV consolidado |
| **Text**    | R (stringr, tidytext), Python  |
| **Graphs**  | igraph, ggraph, tidygraph     |
| **Viz**     | ggplot2, redes por país/facet |

::: {.fragment}
Reproducible pipeline: scripts for consolidation → enrichment → network construction → figures.
:::

## Takeaways

1. **Elite identification** in LATAM can be operationalized with text-derived entities and relations.
2. **Networks** make kinship and institutional ties explicit and comparable across countries.
3. **Text mining** fills gaps when structured data is missing (e.g. nationality, offices, links).
4. **Combining NLP + networks** fits COMPTEXT’s focus on computational approaches to political text and structure.

## Thank you

**Elite identification in Latin America: networks, text mining & NLP**

Matías Deneken · COMPTEXT 2026

::: {.fragment}
Questions and suggestions welcome.
:::
